Too good to be false: Nonsignificant results revisited

### 2017-01-18 submission/manuscript.Rnw
Update ZENODO DOI

\section*{Author note}
All research files, data, and analyses scripts are preserved and made available for download at \url{http://doi.org/10.5281/zenodo.154514}.
\newpage

>>>>

\section*{Author note}
All research files, data, and analyses scripts are preserved and made available for download at \url{http://doi.org/10.5281/zenodo.250492}.
\newpage

### 2017-01-18 submission/manuscript.tex
Update ZENODO DOI

\begin{abstract}
Due to its probabilistic nature, Null Hypothesis Significance Testing (NHST) is subject to decision errors. The concern for false positives has overshadowed the concern for false negatives in the recent debates in psychology. This is unwarranted,  since  reported statistically nonsignificant findings may just be 'too good to be false'. We examined evidence for false negatives in nonsignificant results in three different ways. We adapted the Fisher method to detect the presence of at least one false negative in a set of statistically nonsignificant results. Simulations show that the adapted Fisher method generally is a powerful method to detect false negatives. We examined evidence for false negatives in the psychology literature in three applications of the adapted Fisher method. These applications indicate that (i) the observed effect size distribution of nonsignificant effects exceeds the expected distribution assuming a null-effect, and approximately two out of three (66.7\%) psychology articles reporting nonsignificant results contain evidence for at least one false negative, (ii) nonsignificant results on gender effects contain evidence of true nonzero effects, and (iii) the statistically nonsignificant replications from the Reproducibility Project Psychology (RPP) do not warrant conclusions about the absence or presence of true zero effects underlying these nonsignificant results. We conclude that false negatives deserve more attention in the current debate on statistical practices in psychology. Neglecting effects due to a lack of statistical power can lead to a waste of research resources and stifle the scientific discovery process.
\end{abstract}

\keywords{nonsignificant, underpowered, effect size, Fisher test}

\section*{Author note}
All research files, data, and analyses scripts are preserved and made available for download at \url{http://doi.org/10.5281/zenodo.154514}.
\newpage

@@

@@

First, we compared the observed nonsignificant effect size distribution (computed with observed test results) to the expected nonsignificant effect size distribution under $H_0$. The expected effect size distribution under $H_0$ was approximated using simulation. We first randomly drew an observed test result (with replacement) and subsequently drew a random nonsignificant $p$-value between 0.05 and 1 (i.e., under the distribution of the $H_0$). Based on the drawn $p$-value and the degrees of freedom of the drawn test result, we computed the accompanying test statistic and the corresponding effect size (for details on effect size computation see Appendix B). This procedure was repeated 163,785 times, which is three times the number of observed nonsignificant test results (54,595). The collection of simulated results approximates the expected effect size distribution under $H_0$, assuming independence of test results. We inspected this possible dependency with the intra-class correlation ($ICC$), where $ICC=1$ indicates full dependency and $ICC=0$ indicates full independence. For the set of observed results, the ICC for nonsignificant $p$-values was 0.001, indicating independence of $p$-values within a paper. The resulting, expected effect size distribution was compared to the observed effect size distribution (i) across all journals and (ii) per journal. To test for differences between the expected and observed nonsignificant effect size distributions we applied the Kolmogorov-Smirnov test. This is a non-parametric goodness-of-fit test for equality of distributions, which is based on the maximum absolute deviation between the independent distributions being compared (denoted $D$; \cite{Massey1951-gj}).

@@

As would be expected, we found a higher proportion of articles with evidence of at least one false negative for higher numbers of statistically nonsignificant results ($k$; see Table \ref{tab:tab4}). For instance, 84\% of all papers that report more than 20 nonsignificant results show evidence for false negatives, whereas 57.7\% of all papers with only 1 nonsignificant result show evidence for false negatives. Consequently, we observe that journals with articles containing a higher number of nonsignificant results, such as JPSP, have a higher proportion of articles with evidence of false negatives. This is the result of higher power of the Fisher method when there are more nonsignificant results.

We also checked whether evidence of at least one false negative at the article level changed over time. Figure \ref{fig:fig4} depicts evidence across all articles per year, as a function of year (1985-2013); point size in the figure corresponds to the mean number of nonsignificant results per article (mean $k$) in that year. Interestingly, the proportion of articles with evidence for false negatives decreased from 77\% in 1985 to 55\% in 2013, despite the increase in mean $k$ (from 2.11 in 1985 to 4.52 in 2013). This decreasing proportion of papers with evidence over time cannot be explained by a decrease in sample size over time, as sample size in psychology articles has stayed stable across time (see Figure \ref{fig:fig5}). One (at least partial) explanation of this surprising result is that in the early days researchers primarily reported fewer APA results and used to report relatively more APA results with 'marginally significant' $p$-values (i.e., $p$-values slightly larger than .05), compared to nowadays. This explanation is supported by both a smaller number of reported APA results in the past and the smaller mean reported nonsignificant $p$-value (0.222 in 1985, 0.386 in 2013). We do not know whether these marginally significant $p$-values were interpreted as evidence in favor of a finding (or not) and how these interpretations changed over time. Another potential explanation is that the effect sizes being studied have become smaller over time (mean effect $r$ 0.257 in 1985, 0.187 in 2013), which results in both higher $p$-values over time and lower power of the Fisher test. Using the data at hand, we cannot distinguish between the two explanations.

>>>>

\begin{abstract}
Due to its probabilistic nature, Null Hypothesis Significance Testing (NHST) is subject to decision errors. The concern for false positives has overshadowed the concern for false negatives in the recent debates in psychology. This might be unwarranted,  since  reported statistically nonsignificant findings may just be 'too good to be false'. We examined evidence for false negatives in nonsignificant results in three different ways. We adapted the Fisher method to detect the presence of at least one false negative in a set of statistically nonsignificant results. Simulations show that the adapted Fisher method generally is a powerful method to detect false negatives. We examined evidence for false negatives in the psychology literature in three applications of the adapted Fisher method. These applications indicate that (i) the observed effect size distribution of nonsignificant effects exceeds the expected distribution assuming a null-effect, and approximately two out of three (66.7\%) psychology articles reporting nonsignificant results contain evidence for at least one false negative, (ii) nonsignificant results on gender effects contain evidence of true nonzero effects, and (iii) the statistically nonsignificant replications from the Reproducibility Project Psychology (RPP) do not warrant strong conclusions about the absence or presence of true zero effects underlying these nonsignificant results. We conclude that false negatives deserve more attention in the current debate on statistical practices in psychology. Potentially neglecting effects due to a lack of statistical power can lead to a waste of research resources and stifle the scientific discovery process.
\end{abstract}

\keywords{NHST, reproducibility project, nonsignificant, power, underpowered, effect size, Fisher test, gender}

\section*{Author note}
All research files, data, and analyses scripts are preserved and made available for download at \url{http://doi.org/10.5281/zenodo.250492}.
\newpage

@@

To put the power of the Fisher test into perspective, we can compare its power to reject the null based on one statistically nonsignificant result ($k=1$) with the power of a regular $t$-test to reject the null. If $\eta=.1$, the power of a regular $t$-test equals 0.17, 0.255, 0.467 for sample sizes of 33, 62, 119, respectively; if $\eta$ = .25, power values equal 0.813, 0.998, 1 for these sample sizes. The power values of the regular $t$-test are higher than that of the Fisher test, because the Fisher test does not make use of the more informative statistically significant findings.

@@

First, we compared the observed nonsignificant effect size distribution (computed with observed test results) to the expected nonsignificant effect size distribution under $H_0$. The expected effect size distribution under $H_0$ was approximated using simulation. We first randomly drew an observed test result (with replacement) and subsequently drew a random nonsignificant $p$-value between 0.05 and 1 (i.e., under the distribution of the $H_0$). Based on the drawn $p$-value and the degrees of freedom of the drawn test result, we computed the accompanying test statistic and the corresponding effect size (for details on effect size computation see Appendix B). This procedure was repeated 163,785 times, which is three times the number of observed nonsignificant test results (54,595). The collection of simulated results approximates the expected effect size distribution under $H_0$, assuming independence of test results in the same paper. We inspected this possible dependency with the intra-class correlation ($ICC$), where $ICC=1$ indicates full dependency and $ICC=0$ indicates full independence. For the set of observed results, the ICC for nonsignificant $p$-values was 0.001, indicating independence of $p$-values within a paper (the ICC of the log odds transformed $p$-values was similar, with $ICC=0.00175$ after excluding $p$-values equal to 1 for computational reasons). The resulting, expected effect size distribution was compared to the observed effect size distribution (i) across all journals and (ii) per journal. To test for differences between the expected and observed nonsignificant effect size distributions we applied the Kolmogorov-Smirnov test. This is a non-parametric goodness-of-fit test for equality of distributions, which is based on the maximum absolute deviation between the independent distributions being compared (denoted $D$; \cite{Massey1951-gj}).

@@

As would be expected, we found a higher proportion of articles with evidence of at least one false negative for higher numbers of statistically nonsignificant results ($k$; see Table \ref{tab:tab4}). For instance, 84\% of all papers that report more than 20 nonsignificant results show evidence for false negatives, whereas 57.7\% of all papers with only 1 nonsignificant result show evidence for false negatives. Consequently, we observe that journals with articles containing a higher number of nonsignificant results, such as JPSP, have a higher proportion of articles with evidence of false negatives. This is the result of higher power of the Fisher method when there are more nonsignificant results and does not necessarily reflect that a nonsignificant $p$-value in e.g. JPSP has a higher probability of being a false negative than one in another journal.

We also checked whether evidence of at least one false negative at the article level changed over time. Figure \ref{fig:fig4} depicts evidence across all articles per year, as a function of year (1985-2013); point size in the figure corresponds to the mean number of nonsignificant results per article (mean $k$) in that year. Interestingly, the proportion of articles with evidence for false negatives decreased from 77\% in 1985 to 55\% in 2013, despite the increase in mean $k$ (from 2.11 in 1985 to 4.52 in 2013). This decreasing proportion of papers with evidence over time cannot be explained by a decrease in sample size over time, as sample size in psychology articles has stayed stable across time (see Figure \ref{fig:fig5}; degrees of freedom is a direct proxy of sample size resulting from the sample size minus the number of parameters in the model). One (at least partial) explanation of this surprising result is that in the early days researchers primarily reported fewer APA results and used to report relatively more APA results with 'marginally significant' $p$-values (i.e., $p$-values slightly larger than .05), compared to nowadays. This explanation is supported by both a smaller number of reported APA results in the past and the smaller mean reported nonsignificant $p$-value (0.222 in 1985, 0.386 in 2013). We do not know whether these marginally significant $p$-values were interpreted as evidence in favor of a finding (or not) and how these interpretations changed over time. Another potential explanation is that the effect sizes being studied have become smaller over time (mean correlation effect $r=$ 0.257 in 1985, 0.187 in 2013), which results in both higher $p$-values over time and lower power of the Fisher test. Using the data at hand, we cannot distinguish between the two explanations.
